V4046\,Sgr high-resolution 1.3\,mm continuum ALMA observations: a disc with a thin ring

### 2020-10-06 mnras_template.tex
Add files via upload

% Abstract of the paper
\begin{abstract}
   %%%%context
    V4046 Sgr is a nearby pre-main sequence spectroscopic binary system that has a gas-rich circumbinary disc.
    We report on new $\sim$20\,mas resolution Atacama Large Millimeter Array (ALMA) observations of the radio continuum at 1.3\,mm, and compare with archival SPHERE-IRDIS differential polarised imaging (DPI) of scattered-light at H-band (1.65\,$\micron$) to reach conclusions on the disc structure. We use a parametric a radiative transfer model of the circumbinary disc, computed with \textsc{radmc3d} to interpret the available data, including the spectral-energy-distribution, in terms of a physical model. In addition, we explore decompositions of the imaging data in polar coordinates to better constrain the disc radial structure. 
    We report a two ringed structure for the millimetre-size dust population with a narrow, less than 4.17$\pm$0.94\,au-wide inner ring located at 13.46$\pm$0.43\,au. 
    This ring is the finest reported so far in a gas rich protoplanetary disc. 
    The ring is surrounded by a $\sim$9\,au-wide gap, flanked by a very sharp outer ring starting at 24\,au. 
    %The radial profile of the outer ring exhibits an intriguing break at $\sim$35\,au. 
    The RT model reproduces accurately the available data.
    %We predict a three ringed structure for the small-dust population with a inner ring at $\sim$6\,au hiding under the coronagraph in the SPHERE observations. 
    %The shadow of the secondary reported previously from the scattered-light data is also reproduced by the RT model. 
    %The DPI image reveals strong forward-scattering at 1.65\,$\micron$, which we reproduced with rather large dust grains of no less than 0.4\,$\micron$, hinting at a deficit of very-small-grains. 
    Using the parametric model scale height as an estimated height, we found that the thin inner ring is stable.
    %Deeper ALMA observations may confirm the lopsided 5\,au ring.

@@

Recent observations of circumstellar discs have revolutionised the way we think about protoplanetary discs, but the focus of resolved imaging with the Atacama Large Millimeter Array (ALMA) or with VLT/SPHERE has been mainly been towards the brighter sources. This report is part of a survey of discs around T-Tauri Stars (DARTSS - ref Avenhaus, Perez et al in pret). Here, we present new 1.3\,mm continuum ALMA observation of V4046\,Sgr, which is a remarkable system where it is possible to observe fundamental processes associated with disc evolution.

@@

To analyse the new ALMA data, we aim to reproduce the disc through a parametric model ran with the radiative transfer code \textsc{radmc3d} \citep{Dullemond_2012} so that the model would agree with the ALMA observation, a previous SPHERE-IRDIS observation and the Spectral Energy Distribution (SED). Combining the model representation of the dust density distribution with a decomposition in polar coordinates of the ALMA image we can have better approximations of the radial structure of the disc. It is worth mentioning that previous models of V4046\,Sgr have been made \citep{Ru_z_Rodr_guez_2019, Rosenfeld_2013, 2019ApJ...882..160Q} but with the difference that our new reference observation of radio wavelengths has a sub-arcsecond resolution, allowing us to discover previously unseen underlying substructures of the large-dust grains disc.

@@

Disc structures can be explored by means of comparing mm continuum and scattered light images. These observations allow us to see the distribution of different populations of dust particles, while ALMA traces the millimetre-sized grains settled in the mid-plane of the disc, scattered light accounts for photons reflected off small micron-sized dust at the disc surface layer, high above the mid-plane. By comparing both is easier to determine whether asymmetries in scattered light correspond to surface ripples or deep structures intrinsic to the underlying density distribution.

@@

In this investigation, a previous polarimetric image at 1.65\,$\micron$ is used. The image was taken at the ESO Very Large Telescope with the SPHERE-IRDIS instrument, a complete description of the image and all about data reduction was reported in \citet{Avenhaus_2018}.

For our analysis of the ALMA image we used an image reconstruction strategy, using the non-parametric image synthesis of the \textsc{uvmem} package \citep{2006ApJ...639..951C, 2018A&C....22...16C} it is possible to super-resolve the clean beam, obtaining an effective angular resolution $\sim$2-3 times higher than in the clean image with natural weights. The model obtains this higher resolution through fitting the data using Chi-squared minimisation. Here we adopted a pure $\chi ^2$ model image. The resulting \textsc{uvmem} image shown in Fig. \ref{fig:two} exhibits previously unseen substructure of the disc. This unique high-definition image reveals that the disc features a two-ring structure for the large grain dust with a broad gap between the rings. The wide and brighter outer ring has been observed in previous occasions, but the thin and weak inner ring has never seen before (\citet{Ru_z_Rodr_guez_2019} anticipated its existence). The outer ring has its peak intensity at $\sim$30\,au and reaches out to $\sim$60\,au. Additionally, there is a surface brightness break at $\sim$35\,au, another feature that has never been considered.

@@

The inner ring is surprisingly narrow, and seems to be off-centre relative to the GAIA stellar position (at the origin of coordinates in Fig. \ref{fig:two}). The ring width can be measured in polar coordinates, following the determination of the inner ring centre and orientation. We minimised the dispersion of the disc radial profiles, from 6\,au to 19\,au (aiming for the inner ring) for a PA of 74.6\,$\degr$, an inclination of 33.9\,$\degr$ and a centre at $\Delta \alpha = 9\pm0.05$\,mas $\Delta \delta = 0.1\pm0.04$\,mas, relative to the stars.

Assuming a Gaussian dust inner ring, we obtained the mean width and the location of the ring by finding for each azimuth the Gaussian intensity profile that best describes the ring. This gave us a FWHM of 4.17$\pm$0.94\,au and that it is located at 13.46$\pm$0.43\,au from the stars (See Fig. \ref{fig:polarring}). As the \textsc{uvmem} model image has an effective angular resolution of $\sim$1/3 that of the natural-weighted beam ($0\farcs04 \,\times \, 0\farcs05$), these measurements only gives an upper limit of the real width.

Doing one more time the minimisation of the dispersion of the disc radial profiles, but this time aiming for the outer ring from 20\,au to 70\,au, for a PA of 76.8\,$\degr$, with an inclination of 34.0\,$\degr$ and a centre at $\Delta \alpha = 19\pm0.04$\,mas $\Delta \delta = 7\pm0.03$\,mas relative to the stars, we can compare the measured location of the inner ring according with this parameters versus to the location obtained before. The deviation is shown in Fig. \ref{fig:polarring} and is not trivial since it depend on a few parameters such as the different inclinations of the rings, the difference in position angle and the small shifts of the centres relative to the stars. Those parameters play an important role, but a possible small difference in eccentricities between the inner and the outer ring could also help explain this deviation between the two measurements. ***

@@

The simulated image at 1.65\,$\micron$ shows a similar radial structure to the one visible in the observations displaying a two ringed disc, where the innermost small-dust ring of the parametric model hides under the artificial coronagraph. The visible asymmetry in the SPHERE observations is recreated using grains larger than 0.4\,$\micron$ as smaller grains do not cause a strong forward scattering, meaning that the disc is depleted of very-small grains. Interestingly, the model accurately shows the shadows described by \citet{dOrazi} that are present in the Sphere-IRDIS image.

The simulated 1.3\,mm continuum image displays some clear similarities with the ALMA observation. The model reproduced the two rings that are present in the observation: the faint inner ring and the outer brighter one. %The outer ring has an intensity peak at $\sim$30\,au and a break present at $\sim$35\,au. 
As the radial profiles obtained from the simulated images of the model closely resemble those deduced from the observations (Fig. \ref{fig:radprofiles}), we can assume that the model provides a reliable approximation of the disc structure, including, particularly for our interest, the dimensions of the previously unseen 1.3\,mm thin inner ring. Accordingly, taking the parametric model values for the large-dust inner ring we have that it is located at 13.6\,au, has an estimated width of 2.6\,au, an estimated scale height ranging from 0.25 to 0.32\,au and a dust mass of about 3\,M$_{\earth}$.

As \citet{2018ApJ...869L..46D} explains, the stability of a narrow dust ring can be analysed comparing its height to its radial extent. For a ring to be stable and long-lived its horizontal dimension can not be less than its vertical height. Using the parametric model scale height of 0.28\,au at 13.46\,au and the measured width of the inner ring (4.17\,au, see Section \ref{sec:Observations}), we have that its width is $\sim$15 times its scale height, meaning that it is stable and the disc is capable of supporting a thin ring as this one.

@@

New ALMA 1.3\,mm continuum imaging of the circumbinary disc around V4046\,Sgr reveals new information about its substructure. Using a radiative transfer model, a preexisting scattered light observation and decompositions in polar coordinates of the computed \textsc{uvmem} model image, we measured and analysed the previously unseen thin inner ring and studied the small-dust population distribution.

The key conclusions of our analysis are as follows.
\begin{itemize}
  \item We report a narrow inner ring for the 1.3\,mm continuum located at 13.46$\pm$0.43\,au from the stars and has an estimated width of 4.17$\pm$0.94\,au. The location of this ring is similar to the inner ring observed in the scattered light image, this reveals that the ring includes a considerable mass of millimetre-sized grains of around 3\,M$_{\earth}$. Using the parametric model scale height value ($h= $ 0.28\,au at 13.46\,au) we have that the ring width is $\sim$15 times its estimated height, making it a stable ring.

@@

  \item We interpret the asymmetry observed with SPHERE-IRDIS at 1.65\,$\micron$ due to strong forward-scattering, which implies that the dust population is depleted of grains smaller than $\sim$0.4\,$\micron$.

>>>>

% Abstract of the paper
\begin{abstract}
    %context
    V4046 Sgr is a nearby pre-main sequence spectroscopic binary system that has a gas-rich circumbinary disc.
    %aim
    We report on new $\sim20$\,mas resolution Atacama Large Millimeter Array (ALMA) observations of the radio continuum at 1.3\,mm, and compare with archival SPHERE-IRDIS differential polarised imaging (DPI) of scattered-light at H-band (1.65\,$\micron$) to reach conclusions on the disc structure. 
    %methods
    We use a parametric a radiative transfer model of the circumbinary disc, computed with \textsc{radmc3d} to interpret the available data, including the spectral-energy-distribution, in terms of a physical model. In addition, we explore decompositions of the imaging data in polar coordinates to better constrain the disc radial structure. 
    %results
    We report a two ringed structure for the millimetre-size dust population with a narrow, previously unseen, less than 4.17$\pm$0.94\,au-wide inner ring centred at 13.46$\pm$0.43\,au. This ring is the finest reported so far in a gas rich protoplanetary disc. The ring is surrounded by a $\sim$9\,au-wide gap, flanked by a very sharp outer ring starting at 24\,au. The radial profile of the outer ring exhibits an intriguing discontinuity at $\sim$35\,au. The RT model reproduces accurately the available data. We predict a three ringed structure for the small-dust population with a inner ring at $\sim$6\,au hiding under the coronagraph in the SPHERE observations. The shadow of the secondary reported previously from the scattered-light data is also reproduced by the RT model. 
    %conclusions
    The DPI image reveals strong forward-scattering at 1.65\,$\micron$, which we reproduced with rather large dust grains of no less than 0.4\,$\micron$, hinting at a deficit of very-small-grains. Using the parametric model scale height as an estimated height, we found that the thin inner ring is stable. Deeper ALMA observations may confirm the lopsided 5\,au ring.

@@

Recent observations of circumstellar discs have revolutionised the way we think about protoplanetary discs, but the focus of resolved imaging with the Atacama Large Millimeter Array (ALMA) or with VLT/SPHERE has been mainly been towards the brighter sources. This report is part of a survey of discs around T-Tauri Stars (DARTSS - ref Avenhaus, Perez et al in pret). Here, we present new 1.3\,mm continuum ALMA observation of V4046\,Sgr.

@@

To analyse the new ALMA data, the aim was to reproduce the disc through a parametric model computed with the radiative transfer code \textsc{radmc3d} \citep{Dullemond_2012} and thus be able to interpret the data in terms of a physical model. The model would have to agree not only with the new ALMA observation but also with a previous SPHERE-IRDIS observation and the Spectral Energy Distribution (SED). Combining the model representation of the dust density distribution with a decomposition in polar coordinates of the ALMA image we can have better approximations of the radial structure of the disc. It is worth mentioning that previous models of V4046\,Sgr have been made \citep{Ru_z_Rodr_guez_2019, Rosenfeld_2013, 2019ApJ...882..160Q} but with the difference that our new reference observation of radio wavelengths has a sub-arcsecond resolution, allowing us to discover previously unseen underlying substructures of the large-dust grain disc.

@@

Disc structures can be explored by means of comparing mm continuum and scattered light images. These observations allow us to see the distribution of different populations of dust particles, while ALMA traces the millimetre-sized grains settled in the mid-plane of the disc, scattered light accounts for photons reflected off small micron-sized dust at the disc surface layer, high above the mid-plane. By comparing both, is easier to determine whether asymmetries in scattered light correspond to surface ripples or deep structures intrinsic to the underlying density distribution.

@@

In this investigation a previous polarimetric image at 1.65\,$\micron$ is used. The image was taken at the ESO Very Large Telescope with the SPHERE-IRDIS instrument, a complete description of the image and all about data reduction was reported in \citet{Avenhaus_2018}.

For our analysis of the ALMA image we used an image reconstruction strategy, using the non-parametric image synthesis of the \textsc{uvmem} package \citep{2006ApJ...639..951C, 2018A&C....22...16C} it is possible to super-resolve the clean beam, obtaining an effective angular resolution $\sim$2-3 times higher than in the clean image with natural weights. The model obtains this higher resolution through fitting the data using Chi-squared minimisation. Here we adopted a pure $\chi ^2$ model image. The resulting \textsc{uvmem} image shown in Fig. \ref{fig:two} exhibits previously unseen substructure of the disc. This image reveals that the disc features two rings of large dust grains with a broad gap between them. The wide and brighter outer ring has its peak intensity at $\sim$30\,au, has a surface brightness break at $\sim$35\,au, and reaches out to $\sim$60\,au. The thin and weak inner ring had never seen before, but \citet{Ru_z_Rodr_guez_2019} did anticipate its existence.

@@

The inner ring is surprisingly narrow and seems to be off-centre relative to the GAIA stellar position (at the origin of coordinates in Fig. \ref{fig:two}). The ring width can be measured in polar coordinates following the determination of the inner ring centre and orientation. We minimised the dispersion of the disc radial profiles, from 6\,au to 19\,au (aiming for the inner ring) for a PA of 74.6\,$\degr$, an inclination of 33.9\,$\degr$ and a centre at $\Delta \alpha = 9\pm0.05$\,mas $\Delta \delta = 0.1\pm0.04$\,mas, relative to the stars.

Assuming a Gaussian dust inner ring, we obtained its mean width and location by finding for each azimuth the Gaussian intensity profile that best describes the ring. This gave us a FWHM of 4.17$\pm$0.94\,au and that it is located at 13.46$\pm$0.43\,au from the stars (See Fig. \ref{fig:polarring}). As the \textsc{uvmem} model image has an effective angular resolution of $\sim$1/3 that of the natural-weighted beam ($0\farcs04 \,\times \, 0\farcs05$), these measurements only gives an upper limit of the real width.

Doing one more time the minimisation of the dispersion of the disc radial profiles, but this time aiming for the outer ring from 20\,au to 70\,au, for a PA of 76.8\,$\degr$, with an inclination of 34.0\,$\degr$ and a centre at $\Delta \alpha = 19\pm0.04$\,mas $\Delta \delta = 7\pm0.03$\,mas relative to the stars, we can compare the measured location of the inner ring according with this parameters versus to the location obtained before. The deviation for each azimuth is shown in Fig. \ref{fig:polarring}, and is not trivial since it depend on a few parameters such as the different inclinations of the rings, the difference in position angle and the shift of the centres relative to the stars. But even though those parameters play an important role, a possible small difference in eccentricities between the inner and the outer ring could also help explain this deviation between the two measurements. ***

@@

The simulated image at 1.65\,$\micron$ shows a similar radial structure to the one visible in the observations, displaying a two ringed disc, where the innermost small-dust ring of the parametric model hides under the artificial coronagraph. The visible asymmetry in the SPHERE observations is recreated using grains larger than 0.4\,$\micron$ as smaller grains do not cause a strong forward scattering, meaning that the disc is depleted of very-small grains. Interestingly, the model accurately shows the shadows described by \citet{dOrazi} that are present in the Sphere-IRDIS image.

The simulated 1.3\,mm continuum image displays some clear similarities with the ALMA observation. The model reproduced the two rings that are present in the observation: the faint inner ring and the outer brighter one. As the radial profiles obtained from the simulated images of the model closely resemble those deduced from the observations (Fig. \ref{fig:radprofiles}), we can assume that the model provides a reliable approximation of the disc structure, including, particularly for our interest, the dimensions of the previously unseen 1.3\,mm thin inner ring. Accordingly, taking the parametric model values for the large-dust inner ring we have that it is located at 13.6\,au, has an estimated width of 2.6\,au, an estimated scale height ranging from 0.25 to 0.32\,au and a dust mass of about 3\,M$_{\earth}$. For the outer ring, we have that it has an intensity peak at $\sim$30\,au, a break present at $\sim$35\,au and a mass of $\sim$43\,M$_{\earth}$

As \citet{2018ApJ...869L..46D} explains, the stability of a narrow dust ring can be analysed comparing its height to its radial extent. For a ring to be stable and long-lived its horizontal dimension can not be less than its vertical height. Using the parametric model scale height of 0.28\,au at 13.46\,au and the measured width of the inner ring (4.17\,au, see Section \ref{sec:Observations}), we have that its width is $\sim$15 times its scale height, meaning that it is stable.

@@

New ALMA 1.3\,mm continuum imaging of the circumbinary disc around V4046\,Sgr reveals new information about its substructure. Using a radiative transfer model and decompositions in polar coordinates of the computed \textsc{uvmem} model image, we measured and analysed the previously unseen thin inner ring and studied the small-dust population distribution.

The key conclusions of our analysis are as follows.
\begin{itemize}
  \item We report a narrow inner ring for the 1.3\,mm continuum located at 13.46$\pm$0.43\,au from the stars and has an estimated width of 4.17$\pm$0.94\,au. The location of this ring is similar to the inner ring observed in the scattered light image, revealing that the ring includes a considerable mass of millimetre-sized grains of around 3\,M$_{\earth}$. Using the parametric model scale height value ($h= $ 0.28\,au at 13.46\,au) we have that the ring width is $\sim$15 times its estimated height, making it a stable ring.

@@

  \item We interpret the asymmetry observed with SPHERE-IRDIS at 1.65\,$\micron$ as due to strong forward-scattering, which implies that the dust population is depleted of grains smaller than $\sim$0.4\,$\micron$.
